% Angelicos coros con gozo cantad
% Critical poetry edition
% New version 2016/04/20

\begin{poemtitle}
\poemhead{\worktitle{Angélicos coros con gozo cantad} (Puebla, \circa{1680})}
\poemsource{Anonymous, from musical setting by Antonio de Salazar, Puebla, Convento de la Santísima Trinidad (\signature{MEX-Mcen}{CSG.256})}
\end{poemtitle}

\begin{poemtranslation}
\begin{original}

\StanzaSection{12}[Estribillo]
Angélicos coros &
con gozo cantad &
la gloria a Belén, &
que es \critnote{casa de pan}
  {\term{Bethlehem} means \quoted{House of Bread} in Hebrew; linking Incarnation and Eucharist.}%
  . &
Celestes esferas, &
estrellas y luces, &
bajad, bajad, &
y \critnote{el cielo de la tierra}
  {Perhaps, the sky, as opposed to supernatural Heaven (\term{cielo Empyreo}).} &
de gloria llenad. &
Que solo aquel lugar &
que el mundo desprecia &
de Dios es capaz.
\SectionBreak

\StanzaSection{4}[Coplas]
1. Para nacer Dios hombre, &
escoge este portal, &
que él sólo es digno alcázar &
de tanta majestad. \&

\Stanza{4}
2. No puede en los palacios &
nacer su inmensidad, &
porque Dios sólo cabe &
en \critnote{él de la humildad}
  {The stable or the Christian.}%
  . \&

\Stanza{4}
3. Aquestas \critnote{ruinas}
  {The stone ruins of Christ's stable, as in contemporary paintings; and the contrite Christian.}%
   \ quiere &
porque con caridad &
lo derribado busca, &
quién viene a edificar. \&

\Stanza{4}
4. Naced, Señor divino, &
que la justicia ya &
del cielo está mirando, &
que nace la Verdad. \&
\end{original}

\begin{translation}
\StanzaSection{12}
Angelic choirs, &
joyfully sing &
\quoted{Glory} to Bethlehem, &
the \quoted{House of Bread}. &
Celestial spheres, & 
stars and lights, &
descend, descend, &
and fill earth's heaven &
with glory. &
For only that place &
which the world discounts &
is capacious for God. 
\SectionBreak

\StanzaSection{4}
1. For God to be born as a man &
he chooses this stable, &
for it alone is a worthy palace &
for such great majesty. \&

\Stanza{4}
2. Not in the palaces &
can his immensity be born, &
because there is only room for God &
in the one that is humble. \&

\Stanza{4}
3. He favors these ruins &
because with compassionate love &
he seeks that which is torn down, &
since he comes to build. \&

\Stanza{4}
4. Be born, divine Lord, &
for lo, justice &
is looking down from heaven, &
for the Truth is born. \&
\end{translation}
\end{poemtranslation}