\edsection{Abbreviations}

\begin{multicols}{2}
    \setlength{\parindent}{0pt}
    \begin{tabu} to \linewidth{lZ}
        A. & \term{Alto}, \term{Altus}\\
        Ac. & \term{Acompañamiento}, \term{basso continuo}\\
        B. & \term{Bajo}, \term{Bassus}\\
        Ch. & Chorus\\
        CN & See critical notes\\
        Corr. & Editorial correction\\
        \worktitle{DRAE} & Real Academía Española, \worktitle{Diccionario de la
        lengua española}, 23rd ed.\\
        Imprints & Consensus reading of poetry imprints\\
        Leg. & \term{Legajo} (archival folder)\\
    \end{tabu}

    \begin{tabu} to \linewidth{lZ}
        M., mm. & Measure (bar), measures\\
        N., nn. & Musical note, notes\\
        \worktitle{OED} & \worktitle{Oxford English Dictionary Online} (accessed 2017)\\
        S. & Soprano; Used in part listings for highest voice to distinguish
        Tiple vs. Tenor (e.g., \quoted{SSAT})\\
        Sugg. & Editorial suggestion\\
        T. & Tenor\\
        Ti. & \term{Tiple}: Treble, boy soprano\\
        Ti. I-1 & Chorus 1, First Tiple\\
    \end{tabu}
\end{multicols}


\edsubsection{Pitch and Octave Designations}

This edition uses upper- and lowercase pitch names together with prime symbols
to indicate specific pitches.
These designations map onto Helmholtz octave numbers as follows:

\input{intro/report/octavetable}

\edsubsection{Archival Sigla}

\begin{tabu} to \textwidth{llZ}
    \textsc{Siglum} & \textsc{Country} & \textsc{Archive}\\
    \siglum{E-Bbc} 
    & Spain 
    & Barcelona, Biblioteca de Catalunya
    \\

    \siglum{E-CAN} 
    & Spain 
    & Canet de Mar, Arxiu Parròquia de Sant Pere i Sant Pau de Canet de 
    Mar, Bisbat de Girona, Fons Capella de Música
    \\

    \siglum{E-Mn} 
    & Spain 
    & Madrid, Biblioteca Nacional de España
    \\

    \siglum{E-SE} 
    & Spain 
    & Segovia, Catedral, Archivo Capitular
    \\

    \siglum{MEX-Pc} 
    & Mexico 
    & Puebla, Catedral, Archivo Capitular
    \\
   
    \siglum{MEX-Mcen} 
    & Mexico 
    & Mexico City, CENIDIM 
    \foreignlanguage{spanish}{(Centro Nacional de Investigación, Documentación e
    Información Musical Carlos Chávez)}
    \\

    \siglum{GB-Lbl} 
    & United Kingdom 
    & London, British Library
    \\
\end{tabu}





