\section*{Colophon}

This edition was produced with free and open-source software on Fedora and
Debian Linux operating systems.
The text was typeset with the \LaTeX{} document-preparation system, based on 
the work of Donald Knuth and Leslie Lamport.
The music was typeset with Lilypond, version 2.19, created by Han-Wen Nienhuys
and Jan Nieuwenhuizen.
I am grateful to the hundreds of volunteers who build and maintain these 
systems, and to those who provided specific help in programming this document.

The text typeface is EB Garamond, designed by Georg Duffner, based on 1592 type 
specimens by Claude Garamont (it is available through the \LaTeX{} package
\term{ebgaramond}).
The Spanish CZ metrical symbol used in the scores was traced manually by the
editor in Inkscape from a villancico manuscript by Miguel de Irízar.

The complete source code for this edition is available in a Git repository at
\url{http://www.bitbucket.com/andrewacashner/wlscm-villancicos-I} and
\url{http://www.andrewcashner.com/villancicos/}.
Please use the Bitbucket site to file bug reports, corrections, or requests for
alternate versions (such as transposed scores, keyboard reductions, or
instrumental performing parts).


\section*{Acknowledgments}

Thanks are due first to Janette Tilley, editor of the Web Library of
Seventeenth-Century Music, and the anonymous reviewers, all of whom greatly
improved the clarity and rigor of the edition.

I am grateful to the following people and institutions for access to the 
primary sources on which these editions are based: 
the capitular archive of the Cathedral of Puebla de los Ángeles (P. Francisco 
Vázquez, rector; the Illmo. Sr. Carlos Ordaz, \term{canónigo archivista});
CENIDIM, the Mexican national center for music research in Mexico City;
the Biblioteca de Catalunya in Barcelona;
the parochial archive of the Church of Saints Peter and Paul, Canet de Mar, in 
the Archdiocese of Girona, Barcelona province;
the capitular archive of Segovia Cathedral (P. Bonifacio Bartolomé);
the Biblioteca Nacional in Madrid, and
the British Library in London.

Travel for archival research in Mexico and Spain was funded by 
a Jacob K. Javits Fellowship from the United States Department of Education, 
a Pre-Dissertation Research Fellowship from the Council for European Studies at 
Columbia University, 
a Eugene K. Wolf Research Travel Grant from the American Musicological Society, 
and grants from the Department of Music and the Center for Latin American 
Studies at the University of Chicago.
Other research funding was provided by a Dissertation Completion Fellowship 
from the American Council of Learned Societies and the Mellon Foundation.

I thank my doctoral advisor, Robert Kendrick, and the other members of my 
dissertation committee, Anne Walters Robertson, Martha Feldman, Frederick de 
Armas, and María Gembero-Ustárroz.
I am also grateful to the following people for specific help with the poetry 
and music in this edition:
Stephen Black,
Anita Damjanovic, 
Miguel Martínez, 
Gustavo Mauleón Rodríguez,
James Nemiroff, and
Martha Tenorio.
My wife Ann makes all of this possible and my children Ben and Joy make it
worthwhile.
