\noteshead{Antonio de Salazar, \worktitle{Angélicos coros con gozo cantad}}

\begin{notesources}

    \begin{source}
        \sourcedescription{\signature{MEX-Mcen}{CSG.256}, Manuscript performing
        parts from collection of the Convento de la Santísima Trinidad, Puebla}
        \annotation{A 8 de Navidad/ \oldabbrev{M}{o} Salazar}
        \parts{SSA, SATB, \term{Guión}; B. II specified as \term{Órgano}}
    \end{source}

\end{notesources}

Like Cáseda's \worktitle{Qué música divina}, this piece is in the Colección 
Jesús Sánchez Garza at CENIDIM, from the Convento de la Santísima Trinidad in 
Puebla.%
    \Autocite[123--125]{Cashner:PhD}
The collection includes many works by Salazar, who may have been trained in 
Puebla before becoming chapelmaster at Mexico City Cathedral.%
  \Autocites{Koegel:Salazar}[109--157]{Goldman:Responsory}
The performing parts bear the names of the convent sisters who performed them
(Figure~\ref{Salazar}).
The name added in a later hand to the Tiple I-1 part, \quoted{Belona}, would 
appear to be the same person who performed the Alto in Cáseda's \worktitle{Qué 
música divina}.

\begin{inlinetable}
\begin{tabular}{lll}
    Tiple & I-1 & Madre Andrea, [different hand:] belona\\
    Tiple & I-2 & Madre Assumpsion\\
    Alto & I & Madre Sacramento\\
    Tiple & II & Madre Thomasa, [on verso, different hand:] Alphonsa de 
    \oldabbrev{S}{ta} crus [cruz]\\
    Alto & II & Madre Rosa\\
    Tenor & II & Ynesica Baeza\\
    Bajo & II & Madre Mariana\\
\end{tabular}
\end{inlinetable}

The \term{Guión} part is in a different, less mature, hand than the others. 
It is written in portrait orientation rather than landscape, and includes 
barlines in the C meter sections.
The part may have been recopied at a later date to replace a worn original.

\criticalnotesheader
\begin{criticalnotes}
    54
    & Gn., n. 4
    & E\fl{}
    & E; cf. explicit E\fl{} in Ti. I-2
    \\

    61 
    & Ti. II, n. 1
    & B\na{}
    & B, no accidental but B\fl{} key signature; 
    cf. explicit B\na{} in Ti. I-1
    \\

    61 
    & Gn., n. 3
    & F\sh{}
    & F; cf. explicit F\sh{} in B. II
    \\
    
    67
    & Ti. II, n. 2
    & \pitch{F}{4}
    & \pitch{G}{4}; cf. other voices F (correction)
    \\

    68
    & Guión 
    & Fermata 
    & No fermata
    \\
   
    75--79
    & A. I 
    & derribado busca, viena a edificar
    & lo derribado busca quien viene a edificar; too many syllables for
    the notes (correction)
    \\
\end{criticalnotes}
